---
title: "Covid Project"
author: "Hanjun Lai"
date: "2023-06-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is a project for course 'Data Science as a Field', we includes import, tidy and analyze the COVID19 dataset from the Johns Hopkins Github site. For more detail: https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/

**we will use following packages:**
```{r}
  library(readr)
  library(tidyverse)
  library(lubridate)
  library(magrittr)
  library(dplyr)
  library(ggplot2)
  library(prophet)
```


## Step 1: Import the data
```{r Import the data, echo=TRUE, warning=FALSE}
library(readr)
library(tidyverse)
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_US.csv",
                "time_series_covid19_confirmed_global.csv",
                "time_series_covid19_deaths_US.csv",
                "time_series_covid19_deaths_global.csv",
                "time_series_covid19_recovered_global.csv")
urls <- str_c(url_in,file_names)
  
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_US.csv",
"time_series_covid19_confirmed_global.csv",
"time_series_covid19_deaths_US.csv",
"time_series_covid19_deaths_global.csv",
"time_series_covid19_recovered_global.csv")
urls <- str_c(url_in,file_names)

global_cases <- (read_csv(urls[2]))
global_deaths <- (read_csv(urls[4]))
us_cases <- (read_csv(urls[1]))
us_deaths <- (read_csv(urls[3]))


```
## Step 2: Data Transformation
```{r  Data_Transformation, warning=FALSE}
global_cases <- global_cases %>% 
      pivot_longer(cols= -c('Province/State','Country/Region',Lat, Long),
                             names_to = "date",
                             values_to = "cases") %>%
         select(-c(Lat,Long))
 global_deaths <- global_deaths %>% 
      pivot_longer(cols= -c('Province/State','Country/Region',Lat, Long),
                                  names_to = "date",
                                  values_to = "deaths") %>%
          select(-c(Lat,Long))
 
  global <- global_cases %>% full_join(global_deaths)
  View(global)

  
 us_cases<- us_cases %>%
   pivot_longer(cols = -(UID:Combined_Key),
                names_to = "date",
                values_to = "cases") %>%
   select(Admin2:cases) %>%
   mutate(date = mdy(date)) %>%
   select(-c(Lat,Long_))
 
 us_deaths<- us_deaths %>%
   pivot_longer(cols = -(UID:Population),
                names_to = "date",
                values_to = "deaths") %>%
   select(Admin2:deaths) %>%
   mutate(date = mdy(date)) %>%
   select(-c(Lat,Long_))
 
 us <- us_cases %>% full_join(us_deaths)
 View(us)

 
 global <- global %>% 
   unite("Combined_Key",
         c('Province/State', 'Country/Region'),
         sep = ", ",
         na.rm = TRUE,
         remove = FALSE)
 
uid_lookup_url <-"https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

uid <- read_csv(uid_lookup_url) %>%
    select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

names(uid)[names(uid) == "Province_State"] <- "Province/State"
names(uid)[names(uid) == "Country_Region"] <- "Country/Region"

global <- global %>%
    left_join(uid, by = c("Province/State","Country/Region")) %>%
    select(-c(UID,FIPS)) %>%
    select('Province/State', 'Country/Region', date, cases, deaths, Population, Combined_Key)

names(global)[names(global) == "Country/Region"] <- "Country_Region"
names(global)[names(global) == "Province/State"] <- "Province_State"
```
## Step 3: Analysis with Visualization 
```{r Analysis with Visualization, warning=FALSE }
#Cases and Deaths visualization in California and US.
US_by_state <- us %>% group_by (Province_State, Country_Region, date) %>%
     summarize (cases = sum (cases), deaths = sum (deaths) ,
                Population = sum(Population)) %>%
     mutate (deaths_per_mill = deaths *1000000 / Population) %>%
     select (Province_State,Country_Region,date, cases, deaths, deaths_per_mill, Population) %>%
     ungroup ()

US_totals <- US_by_state %>% group_by (Country_Region, date) %>%
     summarize (cases = sum (cases), deaths = sum (deaths) ,
                Population = sum(Population)) %>%
     mutate (deaths_per_mill = deaths *1000000 / Population) %>%
     select (Country_Region,date, cases, deaths, deaths_per_mill, Population) %>%
     ungroup ()

US_totals %>%
    filter (cases > 0) %>%
    ggplot (aes(x = date, y = cases))+
    geom_line (aes (color = "cases"))+
    geom_point(aes (color = "cases"))+
    geom_line(aes (y = deaths, color = "deaths"))+
    geom_point (aes (y = deaths, color = "deaths"))+
    scale_y_log10() + 
    theme (legend.position="bottom",
                         axis.text.x = element_text (angle = 90)) +
    labs (title = "COVID19 in US", y= NULL)

state <- 'California'
US_by_state %>%
    filter(Province_State == state) %>%
    filter (cases > 0) %>%
    ggplot (aes(x = date, y = cases))+
    geom_line (aes (color = "cases"))+
    geom_point(aes (color = "cases"))+
    geom_line(aes (y = deaths, color = "deaths"))+
    geom_point (aes (y = deaths, color = "deaths"))+
    scale_y_log10() + 
    theme (legend.position="bottom",
           axis.text.x = element_text (angle = 90)) +
    labs (title = str_c("COVID19 in ", state) , y= NULL)
```
**From the visualization, we can see the speed of increase in both global cases and cases in CA are slowing down**<br><br><br><br><br><br><br>

## Visualizion

```{r, Visualizion, warning=FALSE}
US_by_state %>%
    filter(date == max(date)) %>%
    arrange(desc(cases)) %>%
    head(20) %>%
    ggplot(aes(x = reorder(Province_State, -cases), y = cases)) +
    geom_bar(stat = 'identity', fill = 'blue') +
    coord_flip() +
    labs(title = "Top 20 US states by number of confirmed cases",
         x = "State",
         y = "Number of confirmed cases")

```
**California has the most confirm cases in United States, and Minnesota is the least. Considering the population each state has, this number makes a lot sense.** <br><br><br><br>

```{r}
US_by_state %>%
    filter(date == max(date)) %>%
    arrange(desc(cases)) %>%
    head(20) %>%
    ggplot(aes(x = reorder(Province_State, -deaths), y = cases)) +
    geom_bar(stat = 'identity', fill = 'blue') +
    coord_flip() +
    labs(title = "Top 20 US states by number of confirmed cases",
         x = "State",
         y = "Number of confirmed cases")
```
**No surprise, the ranking is similar to the casese data** <br><br><br><br>

```{r}
global %>%
    filter(date == max(date)) %>%
    arrange(desc(cases)) %>%
    head(20) %>%
    ggplot(aes(x = reorder(Country_Region, -cases), y = cases)) +
    geom_bar(stat = 'identity', fill = 'red') +
    coord_flip() +
    labs(title = "Top 20 countries by number of confirmed cases",
         x = "Country",
         y = "Number of confirmed cases")
```
 **US has the most cases, and Poland has the least. Surprisely we see India, which is a country with larger population than US, only have less than half of the cases US has**  <br><br><br><br>
```{r}
global %>%
    filter(date == max(date)) %>%
    arrange(desc(cases)) %>%
    head(20) %>%
    ggplot(aes(x = reorder(Country_Region, -deaths), y = cases)) +
    geom_bar(stat = 'identity', fill = 'red') +
    coord_flip() +
    labs(title = "Top 20 countries by number of confirmed cases",
         x = "Country",
         y = "Number of confirmed cases")
```
 
 
## Creating a model

**For time series forecasting of the COVID-19 data, we can use models such as ARIMA (AutoRegressive Integrated Moving Average), SARIMA (Seasonal ARIMA), and prophet from Facebook. I'll use Prophet because it can handle daily data, missing values, and has good performance in many cases.** <br<br><br><br>
```{r}

library(prophet)

df <- US_totals %>% 
  select(date, deaths) %>%
  rename(ds = date, y = deaths)

m <- prophet(df)

future <- make_future_dataframe(m, periods = 365) # predict the next 365 days
forecast <- predict(m, future)
plot(m, forecast)


#Predicted data in dataframe
future_forecast <- forecast %>% 
  filter(ds > max(df$ds))

print(future_forecast[,c('ds','yhat','yhat_lower','yhat_upper')])


```

**I trains the Prophet model on the number of cases in the US and predicts the number of cases for the next 365 days.And I also printed out the result as well, we can see that with our model prediction, by 2024-03-08, the predicted number of deaths in US is 1231904, upper bounded by 1784270, which is a slower rate of increase than the past 365 days**<br><br><br><br><br><br>

#Conclusion
In this project, We analysis data through looking at the cases in US and global, this will help us understand more about the trend of covid-19 is going on, and these kind of analysis is going to be helpful to CDC to determine is any measure will be necessary for citizens safe, and also to the International infectious diseases organization.There are more insight is worth digging, like correlation between variables, trying to find connection between cases and area, etc.