---
title: "NYPD_shooting_project"
author: "Hanjun Lai"
date: "2023-05-15"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary 

This is a project for course 'Data Science as a Field', a data report base on ' NYPD_Shooting_Incident_Data__Historic_' data. For more detail: https://catalog.data.gov/dataset/nypd-shooting-incident-data-historic 
In the report, we will break it down in some data cleaning, data visualization and data analysis, and we will get insight from the victim patterns.



**we will use following packages:**

```{r}
  library(readr)
  library(tidyverse)
  library(lubridate)
  library(magrittr)
  library(dplyr)
  library(ggplot2)
```
 
## Step 1: Import the data
```{r NYPD_Project}
library(readr)
data_url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
nypd_project_data <- read_csv(data_url)

```



## Step 2: Data Summary and Cleanup <br>
**we will clean up the data for the coming analysis, and then we will use different transformations to create some new data set to run through the analysis that will help we make the conclusion**
```{r Data Summary and Cleanup}

# Drop columns
nypd_project_data <- nypd_project_data %>%
                      select(-c(Lon_Lat,Longitude,Latitude,X_COORD_CD, Y_COORD_CD))

# Identify missing values
missing_vals <- is.na(nypd_project_data)
summary(missing_vals)

# Drop the columns that is more that 50% of the data is missing
nypd_project_data <- nypd_project_data %>%
                      select(-c(LOCATION_DESC, LOC_CLASSFCTN_DESC, LOC_OF_OCCUR_DESC))

# Transforming the date type to format =  '%m/%d/%Y'
nypd_project_data$OCCUR_DATE <- as.Date(nypd_project_data$OCCUR_DATE, format =  '%m/%d/%Y')

# Transforming the some columns to factor type 
nypd_project_data$PERP_SEX <- as.factor(nypd_project_data$PERP_SEX)
nypd_project_data$VIC_SEX <- as.factor(nypd_project_data$VIC_SEX)
nypd_project_data$BORO <- as.factor(nypd_project_data$BORO)
nypd_project_data$PERP_RACE <- as.factor(nypd_project_data$PERP_RACE)
nypd_project_data$VIC_RACE <- as.factor(nypd_project_data$VIC_RACE)
nypd_project_data$PERP_AGE_GROUP <- as.factor(nypd_project_data$PERP_AGE_GROUP)
nypd_project_data$VIC_AGE_GROUP <- as.factor(nypd_project_data$VIC_AGE_GROUP)

# Summary of the data
summary(nypd_project_data)  

```








## Step 3: Data Visualizations and Analysis<br>

```{r Import packages, echo=FALSE, warning=FALSE}
library(lubridate)
library(magrittr)
library(dplyr)
library(ggplot2)
```
<br>

```{r Crime_count_hours, echo=TRUE}
   # Create a dataset of the crime counts by hour
   crime_count_hour <- nypd_project_data %>%
            mutate(hour = lubridate::hour(OCCUR_TIME)) %>%
            group_by(hour) %>%
            summarise(num_crimes_hour = n())
  
   # Create an enhanced bar plot of the crime counts by hour
   ggplot(crime_count_hour, aes(x = reorder(hour, num_crimes_hour), y = num_crimes_hour, fill = num_crimes_hour)) +
      geom_bar(stat = "identity", color = "black") +
      labs(x = "Hour of Day", y = "Number of Crimes", title = "Crime Counts by Hour") +
      theme_minimal() +
      theme(plot.title = element_text(hjust = 0.5, size = 20, face = "bold"),
            axis.title = element_text(size = 14, face = "bold"),
            axis.text.x = element_text(angle = 90, hjust = 1)) +
      scale_fill_gradient(low = "lightblue", high = "darkblue") 
          
```

**We can tell that from 8:00pm---3:00am(+1day), crime number are all above 1500, indicate a crime peak** <br><br><br><br><br><br>

```{r crime_count_age, echo=TRUE}
# Create a dataset of the crime counts by age
crime_count_age <- nypd_project_data %>%
          group_by(VIC_AGE_GROUP) %>%
          summarise(num_crimes_age = n())
  
# Create an enhanced bar plot of the crime counts by age
ggplot(crime_count_age, aes(x = reorder(VIC_AGE_GROUP, num_crimes_age), 
                            y = num_crimes_age, fill = num_crimes_age)) +
    geom_bar(stat = "identity", color = "black") +
    labs(x = "Age Group", y = "Number of Crimes", title = "Crime Counts by Age") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5, size = 20, face = "bold"),
          axis.title = element_text(size = 14, face = "bold"),
          axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_fill_gradient(low = "lightgreen", high = "darkgreen") 
```

**From the graph we can tell that age group of '25-44' has the highest frequency rate of death, and following by '18-24'**<br><br><br><br><br><br><br>

```{r}
# Create a dataset of the crime counts by sex
crime_count_sex <- nypd_project_data %>%
          group_by(VIC_SEX) %>%
          summarise(num_crimes_sex = n())
  
      # Create an enhanced bar plot of the crime counts by sex
ggplot(crime_count_sex, aes(x = reorder(VIC_SEX, num_crimes_sex), 
                            y = num_crimes_sex, fill = num_crimes_sex)) +
    geom_bar(stat = "identity", color = "black") +
    labs(x = "Sex", y = "Number of Crimes", title = "Crime Counts by Sex") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5, size = 20, face = "bold"),
          axis.title = element_text(size = 14, face = "bold")) +
    scale_fill_gradient(low = "pink", high = "red")
```

**Male victims are much more than female victims.**<br><br><br><br><br><br>

```{r}
# Create a dataset of the crime counts by Borough
crime_count_borough <- nypd_project_data %>%
          group_by(BORO) %>%
          summarise(num_crimes_borough = n())
  
     # Create an enhanced bar plot of the crime counts by borough
ggplot(crime_count_borough, aes(x = reorder(BORO, num_crimes_borough), 
                            y = num_crimes_borough, fill = num_crimes_borough)) +
    geom_bar(stat = "identity", color = "black") +
    labs(x = "Borough", y = "Number of Crimes", title = "Crime Counts by Borough") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5, size = 20, face = "bold"),
          axis.title = element_text(size = 14, face = "bold"),
          axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_fill_gradient(low = "lightblue", high = "darkblue") 
```

**Brooklyn has the highest crime rate in NYC**<br><br><br><br><br><br>

```{r Trend by Years, echo=TRUE}
 # Create dataset of the crime counts by year
crime_count_year <- nypd_project_data %>%
                                   mutate(year = lubridate::year(OCCUR_DATE)) %>%
                                   group_by(year) %>%
                                   summarise(num_crimes_year = n())

# Create an enhanced bar plot of the crime counts by year
ggplot(crime_count_year, aes(x = year, 
                             y = num_crimes_year, fill = num_crimes_year)) +
    geom_bar(stat = "identity", color = "black") +
    geom_line(data = crime_count_year, color='red',size=2) +
    labs(x = "Year", y = "Number of Crimes", title = "Crime Counts by Year") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5, size = 20, face = "bold"),
          axis.title = element_text(size = 14, face = "bold")) +
    scale_fill_gradient(low = "lightyellow", high = "darkorange") 

                      
```
**Overall from year 2006 to 2019, the number of crime decrease in long-run, from about 2000 cases in 2006 to less than 1000 cases in 2019.but from 2020, the cases bounce back strong and reaching the record high since 2006.**<br><br><br><br><br><br>

## Step 4: Conclusion and Bias Identification<br><br>


In conclusion, our analysis of the NYPD Shooting Incident Data reveals several noteworthy patterns. It appears that the majority of victims fall within the 25-44 age group, and males are disproportionately represented among victims. Geographically, Brooklyn holds the highest number of reported incidents.

Interestingly, our data suggest a decline in incidents from 2006 to 2019, followed by a surge in 2020, leading to a record high since 2006. It would be helpful to further investigate the factors that contributed to this sudden increase. It is also essential to mention that the night hours (8:00 pm - 3:00 am) are the most active periods for these incidents, which might be a crucial factor to consider in designing preventive strategies.

However, our analysis should be interpreted with caution due to potential biases and limitations in the data. Firstly, the data is entirely dependent on reported incidents, and thus it could be subject to underreporting bias. Not all shooting incidents are reported or correctly recorded, and the chances of an incident being reported might vary by factors such as location, the socioeconomic status of the victim, and the time of the incident.

Secondly, there is a significant amount of missing data for some variables, especially for the sex of the victims. The handling of these missing data, whether by imputation or complete case analysis, could potentially introduce bias into the analysis.

As for personal biases, as a data analyst, I aimed to approach this analysis objectively and avoid allowing personal beliefs or preconceptions to influence the interpretation of the results. I made efforts to present the data as it is, and base conclusions strictly on what the data reveals. It is, however, important to be mindful of potential unconscious biases that could affect the analytical process and to continually question the findings and assumptions.

Moving forward, it would be useful to incorporate additional data, such as socioeconomic indicators or more specific location data, to gain a more comprehensive understanding of the factors contributing to these incidents. Moreover, exploring more sophisticated statistical modeling approaches could help in identifying complex relationships and patterns in the data.

Overall, despite the limitations and potential biases, this analysis provides valuable insights into shooting incidents in New York City and offers a basis for further investigations and decision-making.